---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r data_load, cache=TRUE} 
dataset <- read.csv("activity.csv", sep=",")
dataset_without_na <- dataset[complete.cases(dataset),]
```

Let's see 'summary' and 'head' for dataset with and without NAs

See the 'summary' of dataset has 2304 NAs but 'summary' of dataset_without_NA doesn't have any NAs
```{r data_summary_&_head}
summary(dataset)
summary(dataset_without_na)
head(dataset)
head(dataset_without_na)
```


## What is mean total number of steps taken per day?
Below code is used to calculate total number of steps taken per day
```{r total_steps_per_day}
total_steps_per_day <- tapply(dataset$steps,dataset$date,sum)
head(total_steps_per_day)
hist(total_steps_per_day,25,ylim=c(0,20), main="Total number of steps taken per day", xlab="Total steps")
```


Below code is used to calculate mean and median of total number of steps taken per day
```{r mean_&_median}
mean_steps_per_day <- tapply(dataset$steps,dataset$date,mean)
head(mean_steps_per_day)
median_steps_per_day <- tapply(dataset$steps,dataset$date,median)
head(median_steps_per_day)
```


## What is the average daily activity pattern?
Code for plotting time series plot
```{r average_steps_time_series_plot}
mean_steps_per_interval <- tapply(dataset_without_na$steps,dataset_without_na$interval,mean)
head(mean_steps_per_interval)
m <- cbind(mean_steps_per_interval,dataset[1:288,3])
colnames(m) <- c("Mean.steps","Interval")
head(m)
plot(m[,2],m[,1], type="l", xlab="Interval", ylab="Average number of steps taken")
```


Code for finding interval that contains maximum number of steps
```{r maximum_steps_interval}
max_steps <- max(mean_steps_per_interval)
max_steps
index_of_max_steps <- which(mean_steps_per_interval==max_steps)
interval_of_max_steps <- dataset_without_na[index_of_max_steps,3]
interval_of_max_steps
```
Interval having maximum number of steps is **`r interval_of_max_steps`** which contains *`r max_steps`* steps


## Imputing missing values
Below code calculates total number of missing values
```{r missing_values_count}
number_of_rows_with_NA <- sum(is.na(dataset))  
number_of_rows_with_NA
```
Total number of missing values are **`r number_of_rows_with_NA`**

Strategy --> Mean steps per 5-minute interval is used to replace NAs
```{r NA_replacement, cache=TRUE}
round_off <- round(mean_steps_per_interval)  
v <- cbind(round_off,dataset[1:length(round_off),3])
colnames(v) <- c("Mean.steps","Interval")
head(v)
new_dataset <- dataset
for(i in 1:dim(v)[1]){
        for(j in 1:dim(new_dataset)[1]){
                if(is.na(new_dataset[j,1]) && new_dataset[j,3]==v[i,2]){
                        new_dataset[j,1] <- v[i,1]
                }
        }
}
```

New dataset with replaced NAs is shown below
```{r NA_imputed_data}
head(new_dataset)
```

Below code is used to calculate total number of steps taken per day for new dataset
```{r new_total_steps_per_day}
new_total_steps_per_day <- tapply(new_dataset$steps,new_dataset$date,sum)
head(new_total_steps_per_day)
hist(new_total_steps_per_day,25,ylim=c(0,20), main="Total number of steps taken per day", xlab="Total steps")
```


Below code is used to calculate mean and median of total number of steps taken per day for new dataset
```{r new_mean_&_median}
new_mean_steps_per_day <- tapply(new_dataset$steps,new_dataset$date,mean)
head(new_mean_steps_per_day)
new_median_steps_per_day <- tapply(new_dataset$steps,new_dataset$date,median)
head(new_median_steps_per_day)
```

Below code shows impact of imputing missing values on total steps
```{r total_steps_in_old_&_new_dataset}
total_steps <- sum(dataset_without_na[,1])
total_steps
new_total_steps <- sum(new_total_steps_per_day)
new_total_steps
total_steps_diff <- new_total_steps - total_steps
total_steps_diff
```
'Total steps' difference for old and new dataset is **`r total_steps_diff`**


## Are there differences in activity patterns between weekdays and weekends?
Code to convert date data in our dataset to 'Date' class
```{r date_class_conversion}
new_dataset$date <- as.Date(new_dataset$date)
```

Below code creates new column indicating day type
```{r day_type}
new_dataset$day.type <- weekdays(new_dataset$date, abbreviate=FALSE)
head(new_dataset)
```

Below code converts day type to either weekday or weekend
```{r conversion_to_weekend_or_weekday}
for(i in 1:dim(new_dataset)[1]){                                      
        if(new_dataset[i,4]=="Saturday" | new_dataset[i,4]=="Sunday"){
                new_dataset[i,4] <- "weekend"
        }
        else{new_dataset[i,4] <- "weekday"}
}
head(new_dataset)
```

Code for plotting time series plot
```{r weekend_&_weekday_time_series_plot, fig.height=14}
new_mean_steps_per_interval <- with(new_dataset, tapply(steps, list(interval,day.type),mean))
head(new_mean_steps_per_interval)
k <- cbind(new_dataset[1:length(round_off),3], new_mean_steps_per_interval)
colnames(k) <- c("Interval","weekday.mean.steps","weekend.mean.steps")
head(k)
par(mfrow=c(2,1))
plot(k[,1], k[,3], type="l", xlab="Interval", ylab="Number of steps", main="weekend")
plot(k[,1], k[,2], type="l", xlab="Interval", ylab="Number of steps", main="weekday")
```